Observing Earth as an Exoplanet with \TESS: Photometric Mapping of Surface Features in Reflected Light

### 2019-08-27 tex/ms.tex
finished revisions

We solve Equation~(\ref{eq:model}) for the weights $\mathbf{w}_n$ and $\mathbf{y}$
by maximizing the negative log likelihood function in two separate steps. In the
first step, we take advantage of the linearity of the problem to perform a fast,
semi-analytical optimization to obtain a starting guess for the weights. In the
second step, we apply additional constraints to prevent overfitting and ensure
an albedo in the range $[0, 1]$ everywhere on the surface.

@@

In general, we find that the iterative scheme converges within about 50 
iterations, taking about two minutes on a laptop computer. However, since it 
is strongly regularized, the model significantly
underfits the data. In the second step, we remove our L2 prior on the spherical
harmonic coefficients and instead regularize the actual surface albedo, which
we compute on a uniform spherical grid with 50,000 points. We enforce a 
uniform prior in the range $[0, 1]$, with a small amount of Gaussian smoothing on either 
side. We additionally impose a constraint on the power spectrum of the
inferred map, requiring that it be drawn from a distribution whose mean
is a decaying power law in the spherical harmonic degree $l$ in order
to suppress ringing and spurious high order features.
Our full negative log likelihood function in this step is therefore
%
\begin{align}
    \label{eq:nll}
    -\log\mathcal{L} \, = \, 
        &\frac{1}{2}(\mathbf{f} - \mathbf{m})^\top \boldsymbol{\Sigma}^{-1} (\mathbf{f} - \mathbf{m}) \, + \nonumber \\
        &\frac{1}{2}\mathbf{w}^\top \boldsymbol{\Lambda}_w^{-1} \mathbf{w} \, + \nonumber \\
        &\frac{1}{2}(\mathbf{a_-} + 1 - \mathbf{a_+})^\top \boldsymbol{\Lambda}_a^{-1} (\mathbf{a_-} + 1 - \mathbf{a_+}) \, + \nonumber \\
        &\frac{1}{2}(\boldsymbol{\rho} - \boldsymbol{l}^{-\gamma})^\top \boldsymbol{\Lambda}_\rho^{-1} (\boldsymbol{\rho} - \boldsymbol{l}^{-\gamma}) \, + \nonumber \\
        &\frac{1}{2} \frac{(\gamma - \gamma_0)^2}{\sigma_\gamma^2}
        \quad,
\end{align}
%
where $\mathbf{f}$ is the measured flux across all targets,
$\mathbf{m}$ is the full model (Equation~\ref{eq:model}),
$\boldsymbol{\Sigma}$ is the (diagonal) data covariance, 
$\mathbf{w}$ are the systematics model weights for all targets,
$\boldsymbol{\Lambda}_w$ is the prior covariance of the weights
(Equation~\ref{eq:wprior}), $\mathbf{a_-}$ and $\mathbf{a_+}$ are
vectors containing the values of the albedo that are below zero and above one, 
respectively, $\boldsymbol{\Lambda}_a$ is the covariance of the Gaussian
used to smooth the edges of the top-hat prior on $\mathbf{a}$,
$\boldsymbol{\rho}$ is the power spectrum of the inferred spherical harmonic map
(equal to the sum of the squares of the coefficients at each degree $\boldsymbol{l}$),
$\gamma$ is the power law index of the power spectrum prior, $\boldsymbol{\Lambda}_\rho$ is the
covariance of that prior, $\gamma_0 = 1.5$ is the mean of the prior on $\gamma$
and $\sigma_\gamma^2 = 10^{-2}$ is the corresponding variance.
%
For simplicity, all our prior covariances are diagonal and
homoscedastic with variances equal to $2.5\times 10^{-3}$ for $\mathbf{w}$,
$10^{-8}$ for $\mathbf{a}$, and $10^{-1}$ for $\boldsymbol{\rho}$.

@@

In our analysis, we also neglected the effects of limb darkening. Because of
atmospheric attenuation, features near the limb of the planet should in
reality contribute less to the flux than our model predicts. This could
bias our results particularly in cases where the planet is seen at half
phase (for instance, during days 1326 and 1341 in Figure~\ref{fig:phases}).
In the absence of limb darkening, the brightest regions on the planet are those
close to the sub-stellar point, which is on the limb. Limb darkening should
shift the regions of peak brightness closer to the center of the planet, adding
an east-west bias in the location of the features we identify.

@@

\begin{figure}[p!]
    \begin{centering}
    \includegraphics[width=\linewidth]{figures/map_temporal.pdf}
    \caption{\label{fig:map_temporal}
             The inferred mean map when allowing for slight temporal
             variations in the spherical harmonic coefficients. Compare
             to Figure~\ref{fig:map}. An animated version of this figure
             is available on 
             \href{https://github.com/rodluger/earthshine/blob/master/tex/figures/map_temporal.mp4}{GitHub}.
             \codelink{earthshine_S1_S2}
             }
    \end{centering}
\end{figure}

@@

Additionally, time-resolved observations in multiple bands could
also help with this problem. This was discussed in \cite{Cowan2009},
who showed that multi-band photometry of the Earth enables one to
disentangle the contributions from oceans, continents, and cloud cover.
A robust detection of surface features on an exoplanet in the habitable
zone will likely necessitate both approaches alongside a circumspect
analysis of the uncertainties and degeneracies at play.

@@

\acknowledgements{%
We gratefully acknowledge Jonathan Fraine, Dan Foreman-Mackey, 
David W. Hogg, Ben Montet, Ben Pope, and Wolfgang Kerzendorf for useful conversations. 
Crucial parts of this work were carried out at the 
TESS Data Workshop, hosted by Space Telescope Science Institute, and the 
Building Early Science with TESS Workshop, hosted by the University of Chicago.}
\facility{TESS}
\software{Astropy, Matplotlib, Numpy, \starry \citep{Luger2019}, 
TensorFlow \citep{Abadi2015}, \spiceypy \citep{Annex2019}}

>>>>

\edited{%
Equation~(\ref{eq:model}) is linear in $\mathbf{w}_n$ and in $\mathbf{y}$
individually, but not in both simultaneously. To find the true optimum for
both vectors, we must therefore perform a nonlinear optimization. However, we find that 
the dimensionality of the problem is so large that unless we initialize the 
optimizer with a good initial guess, it is unable to converge on a good solution.
Fortunately, we are able to obtain an initial guess by exploiting the
linearity of the problem and alternating between linear optimization of $\mathbf{w}_n$
and linear optimization of $\mathbf{y}$. Then, in a second step, we run a
nonlinear optimizer, applying additional nonlinear constraints to prevent overfitting and ensure
an albedo in the range $[0, 1]$ everywhere on the surface.
}

@@

In general, we find that the iterative scheme converges within about 50 
iterations, taking about two minutes on a laptop computer. However, since it 
is strongly regularized, the model significantly
underfits the data. In the second step, we remove our L2 prior on the spherical
harmonic coefficients and instead regularize the actual surface albedo, which
we compute on a uniform spherical grid with 50,000 points. We enforce a 
uniform prior in the range $[0, 1]$, with a small amount of Gaussian smoothing on either 
side. 
\edited{%
Note that while the lower bound of this range corresponds to a true physical
constraint---non-negative surface albedo---the upper bound of unity is
somewhat \emph{ad hoc}, since the true normalization of the flux is
unknown. We find that varying this upper bound (or removing it entirely) 
does not significantly change our results, since we only report ``normalized''
albedos; we discuss this further below.
}%

We additionally impose a constraint on the power spectrum of the
inferred map, requiring that it be drawn from a distribution whose mean
is a decaying power law in the spherical harmonic degree $l$ in order
to suppress ringing and spurious high order features.
\edited{%
In practice, however, we find that this constraint is negligible compared
to our prior on the albedo.
}%
Our full negative log likelihood function in this step is therefore
%
\begin{align}
    \label{eq:nll}
    -\log\mathcal{L} \, = \, 
        &\frac{1}{2}(\mathbf{f} - \mathbf{m})^\top \boldsymbol{\Sigma}^{-1} (\mathbf{f} - \mathbf{m}) \, + \nonumber \\
        &\frac{1}{2}\mathbf{w}^\top \boldsymbol{\Lambda}_w^{-1} \mathbf{w} \, + \nonumber \\
        &\frac{1}{2}(\mathbf{a_-} + 1 - \mathbf{a_+})^\top \boldsymbol{\Lambda}_a^{-1} (\mathbf{a_-} + 1 - \mathbf{a_+}) \, + \nonumber \\
        &\frac{1}{2}(\boldsymbol{\rho} - \boldsymbol{l}^{-\gamma})^\top \boldsymbol{\Lambda}_\rho^{-1} (\boldsymbol{\rho} - \boldsymbol{l}^{-\gamma}) \, + \nonumber \\
        &\frac{1}{2} \frac{(\gamma - \gamma_0)^2}{\sigma_\gamma^2}
        \quad,
\end{align}
%
where $\mathbf{f}$ is the measured flux across all targets,
$\mathbf{m}$ is the full model (Equation~\ref{eq:model}),
$\boldsymbol{\Sigma}$ is the (diagonal) data covariance, 
$\mathbf{w}$ are the systematics model weights for all targets,
$\boldsymbol{\Lambda}_w$ is the prior covariance of the weights
(Equation~\ref{eq:wprior}), $\mathbf{a_-}$ and $\mathbf{a_+}$ are
vectors containing the values of the albedo that are below zero and above one, 
respectively, $\boldsymbol{\Lambda}_a$ is the covariance of the Gaussian
used to smooth the edges of the top-hat prior on $\mathbf{a}$,
$\boldsymbol{\rho}$ is the power spectrum of the inferred spherical harmonic map
(equal to the sum of the squares of the coefficients at each degree $\boldsymbol{l}$),
$\gamma$ is the power law index of the power spectrum prior, $\boldsymbol{\Lambda}_\rho$ is the
covariance of that prior, $\gamma_0 = 1.5$ is the mean of the prior on $\gamma$
and $\sigma_\gamma^2 = 10^{-2}$ is the corresponding variance.
%
For simplicity, all our prior covariances are diagonal and
homoscedastic with variances equal to $2.5\times 10^{-3}$ for $\mathbf{w}$,
$10^{-8}$ for $\mathbf{a}$, and $10^{-1}$ for $\boldsymbol{\rho}$.

@@

In our analysis, we also neglected the effects of limb darkening. Because of
atmospheric attenuation, features near the limb of the planet should in
reality contribute less to the flux than our model predicts. This could
bias our results particularly in cases where the planet is seen at half
phase (for instance, during days 1326 and 1341 in Figure~\ref{fig:phases}).
In the absence of limb darkening, the brightest regions on the planet are those
close to the sub-stellar point, which is on the limb. Limb darkening should
shift the regions of peak brightness closer to the center of the planet, adding
an east-west bias in the location of the features we identify.
\edited{%
Similarly, Rayleigh scattering and anisotropic scattering by clouds probably cause 
similar biases, as could specular reflection off the oceans (glint).
}

@@

\begin{figure}[p!]
    \begin{centering}
    \includegraphics[width=\linewidth]{figures/map_temporal.pdf}
    \caption{\label{fig:map_temporal}
             The inferred mean map when allowing for slight temporal
             variations in the spherical harmonic coefficients. Compare
             to Figure~\ref{fig:map}. An 3 second animated version of the 
             top panel is available in the HTML version of this article, showing
             the inferred cloud movement during sectors 1 and 2.
             \codelink{earthshine_S1_S2}
             }
    \end{centering}
\end{figure}

@@

Additionally, time-resolved observations in multiple bands could
also help with this problem. This was discussed in \cite{Cowan2009}%
\edited{, \cite{Kawahara2011}, and \cite{Fujii2012}, all of which showed }%
that multi-band photometry of the Earth enables one to
disentangle the contributions from oceans, continents, and cloud cover.
A robust detection of surface features on an exoplanet in the habitable
zone will likely necessitate both approaches alongside a circumspect
analysis of the uncertainties and degeneracies at play.

@@

\acknowledgements{%
We gratefully acknowledge Jonathan Fraine, Dan Foreman-Mackey, 
David W. Hogg, Ben Montet, Ben Pope, and Wolfgang Kerzendorf for useful conversations. 
Crucial parts of this work were carried out at the 
TESS Data Workshop, hosted by Space Telescope Science Institute, and the 
Building Early Science with TESS Workshop, hosted by the University of Chicago.}
\facility{TESS}
\software{%
\edited{%
Astropy \citep{astropy1,astropy2}, Matplotlib \citep{matplotlib}, Numpy \citep{numpy}, 
}%
\starry \citep{Luger2019}, 
TensorFlow \citep{Abadi2015}, \spiceypy \citep{Annex2019}}
